---
title: "Time Series Forecasting for SFO Air Passengers"
member: "Yan Naing Oo"
date: "`r Sys.Date()`"
output: officedown::rdocx_document
always_allow_html: true
---

# Load Packages

```{r}

library(fpp3)
library(tsibble)
library(readr)
library(tseries)
library(plotly)
library(knitr)
library(officedown)
```

```{r setup, include=FALSE}

knitr::opts_chunk$set(
	echo = FALSE,
	# Table Caption
	tab.cap.style = "Table Caption",
	tab.cap.pre = "Table",
	tab.cap.sep = ": ",
	# Figure Captioning
	fig.cap.style = "Image Caption",
	fig.cap.sep = ": "
	
)
```

# Load data from SFO air passenger .csv file

```{r}
# load the sf air passengers data
air_passenger_data <- read_csv("Air_Traffic_Passenger_Statistics.csv", show_col_types = FALSE) %>%
  mutate(Year =  ymd(paste(substr(`Activity Period`, 1,4), substr(`Activity Period`, 5,6), "01", sep ="/"))) %>%

  rename(
    year = Year,
    operating_airline = `Operating Airline`,
    operating_airline_iata_code = `Operating Airline IATA Code`,
    published_airline = `Published Airline`,
    published_airline_iata_code = `Published Airline IATA Code`,
    geo_summary = `GEO Summary`,
    geo_region = `GEO Region`,
    activity_type_code = `Activity Type Code`,
    price_category_code = `Price Category Code`,
    terminal = `Terminal`,
    boarding_area = `Boarding Area`,
    passenger_count =`Passenger Count`
   
  ) %>%
  
  select(-`Activity Period` )
  glimpse(air_passenger_data)
  
 
```

# Data Exploration

#### Air Passengers Traffic at SFO

```{r}

sfo_passenger <- air_passenger_data %>%
  group_by(year) %>%
  summarize(total = sum(passenger_count))

# Visualize the passenger count over time
plot_ly(data = sfo_passenger,
        x = ~ year,
        y = ~ total/1e6,
        type = "scatter",
        mode = "lines") %>%
  layout(title = "Monthly Air Traffic Passengers at SFO",
         yaxis = list(title = "Passenger (in million)"),
         xaxis = list(title = "Year"))


```

#### Air Passengers Traffic at SFO by Activity Type

```{r}

start <- as.Date("2010-01-01")   
end <- as.Date("2022-12-31")


sfo_passenger_by_activity <- air_passenger_data %>%
  filter(year >= start & year <= end ) %>%
  group_by(year, activity_type_code) %>%
  summarize(total = sum(passenger_count), .groups = "drop")

plot_ly(data = sfo_passenger_by_activity,
        x = ~ year,
        y = ~ total/1e6,
        color = ~ activity_type_code,
        type = "scatter",
        mode = "lines") %>%
  layout(title = "Air Passengers Traffic at SFO by Activity Type",
         yaxis = list(title = "Passenger (in million)"),
         xaxis = list(title = "Year"))


```

#### Air Passengers Traffic at SFO by Geography

```{r}

sfo_passenger_by_geo <- air_passenger_data %>%
  filter(year >= start & year <= end ) %>%
  group_by(year, geo_summary) %>%
  summarize(total = sum(passenger_count), .groups = "drop")

plot_ly(data = sfo_passenger_by_geo,
        x = ~ year,
        y = ~ total/1e6,
        color = ~ geo_summary,
        type = "scatter",
        mode = "lines") %>%
  layout(title = "Air Passengers Traffic at SFO by Geography",
         yaxis = list(title = "Passenger (in million)"),
         xaxis = list(title = "Year"))

sfo_geosummary_pcs <- air_passenger_data %>%
  filter(year >= start & year <= end ) %>%
  group_by(geo_summary) %>%
  summarise(total = sum(passenger_count), .groups = "drop") %>%
  mutate(percent = paste0(round(total / sum(total) * 100, 1), "%"))


```

#### Geo Summary by Passengers

```{r}

ggplot(sfo_geosummary_pcs, aes(x = geo_summary, y = percent, fill = geo_summary)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent), vjust = -0.5) +
  labs(x = "Geo Summary", y = "Percentage", title = "Geo Summary by Passengers") +
  theme_minimal()



```

```{r}

sfo_operating_airline_pcs <- air_passenger_data %>%
  filter(year >= start & year <= end) %>%
  group_by(operating_airline) %>%
  summarise(total = sum(passenger_count/1e6), .groups = "drop") %>%
  mutate(percent = paste0(round(total / sum(total) * 100, 1), "%")) %>%
  arrange(desc(total)) %>%
  top_n(10)

```

#### Top 10 Operating Airlines by Percentage

```{r tab.cap="sfo_top_passenger table", tab.id="sfopassengertab", tab.cap.style="Table Caption"}

head(sfo_operating_airline_pcs,10)


```

# Data Transformation

#### Convert to tisbble format

```{r}

sfo_data <- air_passenger_data %>%
  filter(year >= start & year <= end) %>%
  group_by(year) %>%
  summarise(total = sum(passenger_count/1e6)) 

sfo <- ts(sfo_data$total, start = c(year(sfo_data$year[1]), month(sfo_data$year[1])), frequency = 12) 


sfo_ts <- as_tsibble(sfo) %>%
 # filter(year >= yearmonth("2000") & year <= yearmonth("2022")) %>%
  rename(year = index, total = value) 


```

```{r tab.cap="sfo_passenger table", tab.id="sfotab", tab.cap.style="Table Caption"}

head(sfo_ts)

```

#### Box-Cox transformation for `lambda`

```{r}

# box_cox transformation is stabilizing variance in the time series
lambda <- sfo_ts %>%
  features(total, features = guerrero) %>%
  pull(lambda_guerrero)

bc_lambda = lambda

sfo_ts %>%
  autoplot(box_cox(total, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Monthly SFO Air Passengers with $\\lambda$=",
         round(lambda,2))))

```

Note: We tried implementing Box-Cox transformation, but it did not have much impact on the final results, hence while applying final model fitting we used non transformed data.

#### Decomposition using STL

```{r}

stl_sfo_ts_dcmp <- sfo_ts %>%
  model( STL(total ~ trend(window = 21) + season(window = 13),  robust = TRUE) ) %>%    # default - monthly trend(window=21) & seasonal(window=13) 
  components()

autoplot(stl_sfo_ts_dcmp) +
  labs(title =
    "Decomposition of SFO Air Passengers using STL")

```

#### KPSS test for differencing

```{r}

sfo_ts %>%
features(total, unitroot_kpss)

sfo_ts %>%
features(total, unitroot_ndiffs)

```

p-value (0.02) is significant. Reject null hypothesis. It indicate that the series is not stationary. Therefore a number difference required to make a time series stationary.

#### Dickey-Fuller test

```{r}
# Check stationarity  ?
adf_test <- adf.test(sfo_ts$total, alternative = "stationary")

head(adf_test$p.value)
# p.value > 0.05 then the series is likely non-stationary

```

#### Split train and test set

```{r}

# create test set by 12 month
test <- sfo_ts %>%
  tail(12)
  
# create train set
train <- sfo_ts %>% 
  head(-12)


```

```{r}


# check for train set
min_train <- min(train$year)
max_train <- max(train$year)

print(paste("Train Min: ", min_train, " Train Max: ", max_train))

```

```{r}

# check for test set
min_test <- min(test$year)
max_test <- max(test$year)

print(paste("Test Min: ", min_test, " Test Max: ", max_test))

```

#### Examine ACF and PACF plots for the differenced

```{r}

train %>%
  gg_tsdisplay(difference(total), plot_type='partial') +
  labs(title="Non Seaonal differenced", y="")

train |>
  gg_tsdisplay(difference(total, 12),
               plot_type='partial', lag=36) +
  labs(title="Seasonally differenced", y="")

#train %>%
  #gg_tsdisplay(difference(total,12) %>% difference(), plot_type='partial', lag=36) +
  #labs(title = "Double Seasonally differenced", y="")

  
```

# Model Fitting

#### Fit the models (ets, arima)

```{r}

sfo_fit <- train %>%
  model(
  
     # Exponential smoothing model (ternd + seasonal) no need transformation
    Ets_auto = ETS(total),   
    Ets_ses = ETS(total ~ error("A") + trend("A") + season("N")), 
    Ets_hw_mul = ETS(total ~ error("M") + trend("A") + season("M")),
    Ets_damped_add = ETS(total  ~ error("A") + trend("Ad") + season("A")),
    Ets_damped_mul = ETS(total ~ error("A") + trend("Ad") + season("M")),
    
    # Arima models
    Arima_stepwise = ARIMA(total),
    Arima_search = ARIMA(total, stepwise = FALSE),
    Arima_311 = ARIMA(total ~ pdq(3,1,1)),
    Arima_410 = ARIMA(total ~ pdq(4,1,0)),
    Arima_012 = ARIMA(total ~ pdq(0,1,2)),
    
    Arima012011 = ARIMA(total ~ pdq(0,1,2) + PDQ(0,1,1)),
    Arima210011 = ARIMA(total ~ pdq(2,1,0) + PDQ(0,1,1)),
    Arima011011 = ARIMA(total ~ pdq(0,1,1) + PDQ(0,1,1)),
    Arima212011 = ARIMA(total ~ pdq(2,1,2) + PDQ(0,1,1)),
    Arima210111 = ARIMA(total ~ pdq(2,1,0) + PDQ(1,1,1))
   
   
  )

# forecast (36 months) 3 years in the future
sfo_fc <- sfo_fit %>%
  forecast(h = 36)


sfo_fit %>% pivot_longer(everything(), names_to = "Model name",
                     values_to = "Orders")

```

# Model Evaluation Metrics

#### Accuracy measures of the forecast

```{r}


# calculate the accuracy measures on test set
accuracy(sfo_fc, test) %>%
#accuracy(sfo_fit) %>%
  arrange(RMSE) %>%
  select(.model, .type, RMSE, MAE, MAPE,)


```

-   RMSE (Root Mean Squared Error): measures the average difference between the forecasted and the actual values, taking into account the squared differences.

-   The `lowest RMSE` indicates `better accuracy`.

#### Summary report for fitted time series models

```{r}

selected_models <- sfo_fit %>%
  select(Ets_ses, Ets_damped_mul, Arima012011, Arima011011, Arima210011)

report(selected_models) %>% 
  arrange(AICc) %>%
  select(.model, AIC, AICc, BIC,)

```

-   AICc (Akaike Information Criterion corrected): a `lowest AICc` value indicates a `best-fitting model`.

```{r}

refit <- train %>%
  model(
    Arima011011 = ARIMA(total ~ pdq(0,1,1) + PDQ(0,1,1)),
    Arima210011 = ARIMA(total ~ pdq(2,1,0) + PDQ(0,1,1)),
    Arima012011 = ARIMA(total ~ pdq(0,1,2) + PDQ(0,1,1)),
    Ets_damped_mul = ETS(total ~ error("A") + trend("Ad") + season("M")),
    Ets_ses = ETS(total ~ error("A") + trend("A") + season("N"))
    
    
  )

refc <- refit %>% forecast(h = 36)

train %>%
  autoplot(total) +
  geom_line(aes(y = .fitted, color = .model), data = fitted(refit)) +
  labs(y = "Passengers (in millions)", title="Best 5 fitted models") +
  autolayer(refc)

```

```{r}

gg_tsresiduals(sfo_fit %>% select(Arima011011), lag_max = 36) + labs(title="Residuals Arima011011 model")
gg_tsresiduals(sfo_fit %>% select(Arima012011), lag_max = 36) + labs(title="Residuals Arima012011 model")
gg_tsresiduals(sfo_fit %>% select(Arima210011), lag_max = 36) + labs(title="Residuals Arima210011 model") 
gg_tsresiduals(sfo_fit %>% select(Ets_damped_mul), lag_max = 36) + labs(title="Residuals Ets_damped_mul model") 
gg_tsresiduals(sfo_fit %>% select(Ets_ses), lag_max = 36) + labs(title="Residuals Ets_ses model")


```

#### Ljung-Box test

```{r}

# find degree of freedom 
sfo_fit %>%
select(Arima012011) %>%
report()

augment(sfo_fit) %>%
  filter(.model=="Arima012011") %>%
  features(.innov, ljung_box, lag = 10, dof = 3)


```

For non-seasonal component, ARIMA(0,1,2) has 2 MA terms, for seasonal ARIMA(0,1,1) has 0 MA term estimated. Therefore the degree of freedom should be 5.

Since the p-value \> 5%, not reject H0 at a significant level. The residuals could be considered white noise series and the model has adequately captured the underlying data.

#### Model selection and forecast

```{r}

sfo_fit %>%
  select(Ets_ses, Ets_damped_mul, Arima011011, Arima012011, Arima210011) %>%
  forecast(h=36) %>%
  autoplot(sfo_ts , level=NULL) +
  labs(title = "Forecasted Models",
  y="Passengers (in millions)")

```

```{r}

sfo_fit %>%
  select(Arima012011) %>%
  forecast(h=36) %>%
  autoplot(sfo_ts , level=NULL) +
  labs(title = "Forecasted result for selected model(Arima012011)",
  y="Passengers (in millions)")

```

```{r}

forecasts <- sfo_fit %>%
  select(Arima012011) %>%
  forecast(h = 36)
# Extract forecasted values
  forecast_values <- forecasts$.mean
# Print the forecasted values for model
  print(forecast_values) # average - 6 mils passengers in the 3 year in the forecast.

```

#### Inverse root test

```{r}

gg_arma(sfo_fit %>%
  select(Arima012011))
```

This inverse toot test implies that the model is stable and a good fit to the data. Also the model has capturing the underlying patterns of the time series.
